Constructing Quotient Inductive-Inductive Types

### 2018-10-30 paper.tex
final version (Thorsten needs to check)

Second, the focus of the current work differs. Cartmell's main result
was the establishment of contextual categories as classifying
categories of GATs. In contrast, we do not discuss classifying
categories\footnote{This could be included in future work.}. Instead,
we present more constructions involving algebras; these include an
explicit initial algebra construction and the CwF model, which yields
a small internal type theory of algebras for each signature, and which
in particular allows us to exactly compute eliminators and
homomorphisms as types in the meta type theory.\footnote{This is
demoed in \cite{kaposi_et_al:LIPIcs:2018:9190}, by a Haskell program which
takes a signature as input, type checks it, then outputs the
corresponding induction methods and eliminators as an Agda file.}

\cite{altenkirch2018quotient} also concerns QIITs. There, signatures
are given by a scheme which builds up complete categories of algebras
from lists of functors. This notion of signature is more semantic than
our one, with more overhead in encoding signatures. It also does not
enforce strict positivity, hence it does not always support initial
algebras. Nonetheless, it is shown that well-behaved QIIT signatures
are covered by this scheme. Also, induction is shown to be equivalent
to initiality, but the notion of induction is given only up to
isomorphisms of algebras, in contrast to the current work, where it is
computed strictly.

@@

@@

Given two $\Gamma$-algebras $\gamma,\gamma':\Gamma^\A$, we define the
notion of homomorphism between them using a variant of the logical
relation interpretation for dependent types
\cite{bernardy12parametricity,atkey}. Contexts become binary
relations, types become heterogeneous binary relations indexed over a
relation for a context, and substitutions and terms are interpreted as
the fundamental theorems for the logical relation.
\begin{alignat*}{5}
  & (\Gamma:\Con)^\M && : \Gamma^\A\ra\Gamma^\A\ra\Set \\
  & (A:\Ty\,\Gamma)^\M && : \Gamma^\M\,\gamma^{\mathit{0}}\,\gamma^{\mathit{1}}\ra A^\A\,\gamma^{\mathit{0}}\ra A^\A\,\gamma^{\mathit{1}}\ra\Set \\
  & (\sigma:\Tms\,\Gamma\,\Delta)^\M && : \Gamma^\M\,\gamma^{\mathit{0}}\,\gamma^{\mathit{1}}\ra\Delta^\M\,(\sigma^\A\,\gamma^{\mathit{0}})\,(\sigma^\A\,\gamma^{\mathit{1}}) \\
  & (t:\Tm\,\Gamma\,A)^\M && : (\gamma^M:\Gamma^\M\,\gamma^{\mathit{0}}\,\gamma^{\mathit{1}})\ra A^\M\,\gamma^M\,(t^\A\,\gamma^{\mathit{0}})\,(t^\A\,\gamma^{\mathit{1}})
\end{alignat*}
However, the usual logical relation interpretation does not produce
homomorphisms. The motivation of Reynolds \cite{reynolds83abstraction}
to replace homomorphisms with logical relations was that homomorphisms
do not work for higher order functions. In our case there are no
higher-order functions because the function space $\Pi$ is strictly
positive. Thus, we are able to interpret the universe by function
space instead of relation space. Similarly, the relation for $\El$ is
the graph of the corresponding function. Below we list the
differences: the left hand side ($\blank^{\M'}$) is the usual logical
relation interpretation, the right hand side is the one we use.
\begin{alignat*}{10}
  & \U^{\M'}\,\gamma^{M'}\,T^{\mathit{0}}\,T^{\mathit{1}} && :\equiv T^{\mathit{0}}\ra T^{\mathit{1}}\ra\U  && \U^\M\,\gamma^M\,T^{\mathit{0}}\,T^{\mathit{1}} && :\equiv T^{\mathit{0}}\ra T^{\mathit{1}} \\
  & (\El\,a)^{\M'}\,\gamma^{M'}\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}} && :\equiv a^{\M'}\,\gamma^{M'}\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}} && (\El\,a)^\M\,\gamma^M\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}} && :\equiv a^\M\,\gamma^M\,\alpha^{\mathit{0}} = \alpha^{\mathit{1}} \\
  & (\Pi\,a\,B)^{\M'}\,\gamma^{M'}\,f^{\mathit{0}}\,f^{\mathit{1}} && :\equiv (\alpha^{M'}:a^{M'}\,\gamma^{M'}\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}})\ra\hspace{2em}  && (\Pi\,a\,B)^\M\,\gamma^M\,f^{\mathit{0}}\,f^{\mathit{1}} && :\equiv (\alpha^{\mathit{0}}:a^\A\,\gamma^{\mathit{0}})\ra \\
  &  && \hspace{1.5em} B^{\M'}\,(\gamma^{M'}, \alpha^{M'})\,(f^{\mathit{0}}\,\alpha^{\mathit{0}}) &&  && \hspace{2em} B^\M\,(\gamma^M, \refl)\,(f^{\mathit{0}}\,\alpha^{\mathit{0}}) \\
  &  && \hspace{1.5em} \hphantom{B^{\M'}\,(\gamma^{M'}, \alpha^{M'})\,} (f^{\mathit{1}}\,\alpha^{\mathit{1}}) &&  && \hspace{3.4em}(f^{\mathit{1}}\,(a^\M\,\gamma^M\,\alpha^{\mathit{0}})) \\
  & (t\app u)^{\M'}\,\gamma^{M'} && :\equiv t^{\M'}\,(u^{\M'}\,\gamma^{M'}) && (t\app u)^\M\,\gamma^M && :\equiv \J\,(t^\M\,\gamma^M\,(u^\A\,\gamma))\,(u^\M\,\gamma^M)
\end{alignat*}
For $\Pi$ types, we could use the usual interpretation, but we have a
better choice. E.g. for the successor constructor of the natural
numbers, the original formulation would give the condition
$(n^{\mathit{0}}:Nat^{\mathit{0}})(n^{\mathit{1}}:Nat^{\mathit{1}})(n^{M}:Nat^M\,n^{\mathit{0}}
= n^{\mathit{1}})\ra
Nat^M\,(\mathit{suc}^{\mathit{0}}\,n^{\mathit{0}}) =
\mathit{suc^{\mathit{1}}}\, n^{\mathit{1}}$. The right hand side
variant strictifies this and results in
$(n^{\mathit{0}}:Nat^{\mathit{0}})\ra
Nat^M\,(\mathit{suc}^{\mathit{1}}\,n^{\mathit{0}}) =
\mathit{suc}^{\mathit{1}}\,(Nat^M\,n^{\mathit{0}})$. The price we have
to pay is that interpreting application requires usage of $\J$.

>>>>

@@

\cite{altenkirch2018quotient} also concerns QIITs. There, signatures
are given by a scheme which builds up complete categories of algebras
from lists of functors. This notion of signature is more semantic than
our one, with more overhead in encoding signatures. It also does not
enforce strict positivity, hence the paper does not construct initial algebras.
Nonetheless, it is shown that well-behaved QIIT signatures
are covered by this scheme. Also, induction is shown to be equivalent
to initiality, but the notion of induction is given only up to
isomorphisms of algebras, in contrast to the current work, where it is
computed strictly.

Second, the focus of the current work differs. Cartmell's main result
was the establishment of contextual categories as classifying
categories of GATs. In contrast, we do not discuss classifying
categories. Instead,
we present more constructions involving algebras; these include an
explicit initial algebra construction and the CwF model, which yields
a small internal type theory of algebras for each signature, and which
in particular allows us to exactly compute eliminators and
homomorphisms as types in the meta type theory.\footnote{This is
demoed in \cite{kaposi_et_al:LIPIcs:2018:9190}, by a Haskell program which
takes a signature as input, type checks it, then outputs the
corresponding induction methods and eliminators as an Agda file.}

@@

Given two $\Gamma$-algebras $\gamma,\gamma':\Gamma^\A$, we define the
notion of homomorphism between them using a variant of the logical
relation interpretation for dependent types
\cite{bernardy12parametricity,atkey}. Contexts become binary
relations, types become heterogeneous binary relations indexed over a
relation for a context, and substitutions and terms are interpreted as
the fundamental theorems for the logical relation.
\begin{alignat*}{5}
  & (\Gamma:\Con)^\M && : \Gamma^\A\ra\Gamma^\A\ra\Set \\
  & (A:\Ty\,\Gamma)^\M && : \Gamma^\M\,\gamma^{\mathit{0}}\,\gamma^{\mathit{1}}\ra A^\A\,\gamma^{\mathit{0}}\ra A^\A\,\gamma^{\mathit{1}}\ra\Set \\
  & (\sigma:\Tms\,\Gamma\,\Delta)^\M && : \Gamma^\M\,\gamma^{\mathit{0}}\,\gamma^{\mathit{1}}\ra\Delta^\M\,(\sigma^\A\,\gamma^{\mathit{0}})\,(\sigma^\A\,\gamma^{\mathit{1}}) \\
  & (t:\Tm\,\Gamma\,A)^\M && : (\gamma^M:\Gamma^\M\,\gamma^{\mathit{0}}\,\gamma^{\mathit{1}})\ra A^\M\,\gamma^M\,(t^\A\,\gamma^{\mathit{0}})\,(t^\A\,\gamma^{\mathit{1}})
\end{alignat*}
However, the usual logical relation interpretation does not produce
homomorphisms. The motivation of Reynolds \cite{reynolds83abstraction}
to replace homomorphisms with logical relations was that homomorphisms
do not work for higher order functions. In our case there are no
higher-order functions because the function space $\Pi$ is strictly
positive. Thus, we are able to interpret the universe by function
space instead of relation space. Similarly, the relation for $\El$ is
the graph of the corresponding function. Below we list the
differences: the left hand side ($\blank^{\M'}$) is the usual logical
relation interpretation, the right hand side is the one we use.
\begin{alignat*}{10}
  & \U^{\M'}\,\gamma^{M'}\,T^{\mathit{0}}\,T^{\mathit{1}} && :\equiv T^{\mathit{0}}\ra T^{\mathit{1}}\ra\U  && \U^\M\,\gamma^M\,T^{\mathit{0}}\,T^{\mathit{1}} && :\equiv T^{\mathit{0}}\ra T^{\mathit{1}} \\
  & (\El\,a)^{\M'}\,\gamma^{M'}\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}} && :\equiv a^{\M'}\,\gamma^{M'}\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}} && (\El\,a)^\M\,\gamma^M\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}} && :\equiv a^\M\,\gamma^M\,\alpha^{\mathit{0}} = \alpha^{\mathit{1}} \\
  & (\Pi\,a\,B)^{\M'}\,\gamma^{M'}\,f^{\mathit{0}}\,f^{\mathit{1}} && :\equiv (\alpha^{M'}:a^{M'}\,\gamma^{M'}\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}})\ra\hspace{1.5em}  && (\Pi\,a\,B)^\M\,\gamma^M\,f^{\mathit{0}}\,f^{\mathit{1}} && :\equiv (\alpha^{\mathit{0}}:a^\A\,\gamma^{\mathit{0}})\ra \\
  &  && \hspace{1.5em} B^{\M'}\,(\gamma^{M'}, \alpha^{M'})\,(f^{\mathit{0}}\,\alpha^{\mathit{0}}) &&  && \hspace{2em} B^\M\,(\gamma^M, \refl)\,(f^{\mathit{0}}\,\alpha^{\mathit{0}}) \\
  &  && \hspace{1.5em} \hphantom{B^{\M'}\,(\gamma^{M'}, \alpha^{M'})\,} (f^{\mathit{1}}\,\alpha^{\mathit{1}}) &&  && \hspace{3.4em}(f^{\mathit{1}}\,(a^\M\,\gamma^M\,\alpha^{\mathit{0}})) \\
  & (t\app u)^{\M'}\,\gamma^{M'} && :\equiv t^{\M'}\,(u^{\M'}\,\gamma^{M'}) && (t\app u)^\M\,\gamma^M && :\equiv \J\,(t^\M\,\gamma^M\,(u^\A\,\gamma))\,(u^\M\,\gamma^M)
\end{alignat*}
For $\Pi$ types, we could use the usual interpretation, but we have a
better choice. E.g. for the successor constructor of the natural
numbers, the original formulation would give the condition
$(n^{\mathit{0}}:Nat^{\mathit{0}})(n^{\mathit{1}}:Nat^{\mathit{1}})(n^{M}:Nat^M\,n^{\mathit{0}}
= n^{\mathit{1}})\ra
Nat^M\,(\mathit{suc}^{\mathit{0}}\,n^{\mathit{0}}) =
\mathit{suc^{\mathit{1}}}\, n^{\mathit{1}}$. The right hand side
variant strictifies this and results in
$(n^{\mathit{0}}:Nat^{\mathit{0}})\ra
Nat^M\,(\mathit{suc}^{\mathit{1}}\,n^{\mathit{0}}) =
\mathit{suc}^{\mathit{1}}\,(Nat^M\,n^{\mathit{0}})$. The price we have
to pay is that interpreting application requires usage of $\J$.
